Detecting Causal Structure in Time Series Data

### 2020-11-23 report/report.tex
Fix incorrect formula for null hypothesis rejection
Beautify ABP

By the definition of the chi-squared distribution with one degree of freedom. To reject the null hypothesis $H_0$ at the $100(1-\alpha)\%$ confidence level, we require that $P(T_n = t_n | H_0) < \alpha$. In other words, a low probability of observing this result under the null hypothesis.

@@

We now complicate things by introducing a second coin, with throws denoted by $\{Y_i\}_{i=1}^{m}$ where we assume each throw is i.i.d Bernoulli with parameter $q$ ($Y_i \sim Bern(q)$). Note that the population sizes $n$ and $m$ may be different. We define $L \coloneqq \sum_{i=1}^{m} Y_i$ which is the analogue of $K$. We now set up our hypotheses to be:
%
\begin{align*}
H_0:& \quad p = q \\
H_1:& \quad p \neq q
\end{align*}
%
Proceeding as before we can derive a formula for the test statistic. This time we denote the test statistic by $t_N$ where $N \coloneqq n + m$.
%
\begin{align*}
t_N &\coloneqq \log \frac{\lik(H_1)}{\lik(H_0)}
= \log \frac{\max_{p,q} P(K=k|p) P(L=l|q)}{\max_p P(K=k|p) P(L=l|q=p)} \\
&= \log \frac
{\binom{n}{k} \hat{p}^k (1 - \hat{p})^{n-k} \binom{m}{l} \hat{q}^k (1 - \hat{q})^{m-l}}
{\binom{n}{k} \hat{r}^k (1 - \hat{r})^{n-k} \binom{m}{l} \hat{r}^l (1 - \hat{r})^{m-l}} \\
&= \log \frac
{ \hat{p}^k (1 - \hat{p})^{n-k}}
{ \hat{r}^k (1 - \hat{r})^{n-k}}
+ \log \frac{\hat{q}^k (1 - \hat{q})^{m-l}}{ \hat{r}^l (1 - \hat{r})^{m-l}}
\end{align*} 
%
Where, $\hat{p} \coloneqq k/n, \hat{q} \coloneqq l/m, \hat{r} \coloneqq (k+l)/(n+m) = (n\hat{p} + m\hat{q})/(n+m)$. Using the same tricks as before, we can express this in terms of the chi-squared distance between the various parameters:
%
\begin{align*}
t_N &= \frac{n}{2} \chi^2\left(Bern(\hat{p}||\hat{r})\right)
+ \frac{m}{2} \chi^2\left(Bern(\hat{q}||\hat{r})\right)
+ \epsilon \\
&\approx \frac{1}{2\hat{r}(1-\hat{r})} \left(
n(\hat{p} - \hat{r})^2 + m(\hat{q} - \hat{r})^2
\right) \\
&= \frac{1}{2\hat{r}(1-\hat{r})} \left(
n\left( \frac{m(\hat{p} - \hat{q})}{n+m}\right)^2 + 
m\left( \frac{n(\hat{q} - \hat{p})}{n+m}\right)^2
\right) \\
&= \frac{nm(\hat{p} - \hat{q})^2}{2\hat{r}(1-\hat{r})(n+m)} \\
&= \frac{1}{2} \left( \sqrt{\frac{nm}{\hat{r}(1-\hat{r})(n+m)}} \left(\hat{p} - \hat{q}\right) \right)^2
\end{align*}
%
Under the null hypothesis $H_0$, we require $p=q(=\mu)$; we introduce this third variable $\mu$ to refer to the true mean to avoid ambiguity. Applying the central limit theorem (for sufficiently large $n$ and $m$) and combining Gaussians in the standard way, we have that:
%
\begin{equation*}
\begin{gathered}
\hat{P} \sim \Gaussian \left( \mu, \frac{\mu(1-\mu)}{n} \right) \\
\hat{Q} \sim \Gaussian \left(\mu, \frac{\mu(1-\mu)}{m} \right) \\
\hat{R} \sim \Gaussian \left(\mu, \frac{\mu(1-\mu)}{n+m} \right) \\
\therefore  \sqrt{\frac{nm}{\mu(1-\mu)(n+m)}} (\hat{P} - \hat{Q}) = Z \sim \Gaussian \left(0, 1 \right) \\
\delta \hat{R} \coloneqq \hat{R} - \mu \sim \Gaussian \left( 0, \frac{\mu(1-\mu)}{n+m} \right)
\end{gathered}
\end{equation*}
%
We almost have $T_n$ we just need to demonstrate that $\hat{R}(1-\hat{R})$ is sufficiently close to $\mu(1-\mu)$ for our purposes.
%
\begin{align*}
\hat{R}(1-\hat{R}) &= (\mu + \delta\hat{R})(1-(\mu + \delta \hat{R})) \\
&= \mu(1-\mu) + O(\delta\hat{R}) \\
\therefore \frac{1}{\hat{R}(1-\hat{R})} &= \frac{1}{\mu(1-\mu)} \left(\frac{1}{1+O(\delta\hat{R})}\right) \\
&= \frac{1}{\mu(1-\mu)} (1 + O(\delta\hat{R})) \\
&\approx \frac{1}{\mu(1-\mu)}
\end{align*}
%
We can neglect the terms of order $\delta\hat{R}$ and higher powers, as it is zero mean and for sufficiently large $n+m$ the variance approaches 0. Therefore, we have the desired expression for $T_N$.
%
\begin{equation}
T_N \approx \frac{1}{2} \left(
\sqrt{\frac{nm}{\mu(1-\mu)(n+m)}}
(\hat{P} - \hat{Q})
\right)^2 = \frac{1}{2} Z^2 \sim \frac{1}{2} \chi^2_1
\end{equation}
%
For this test we can also use the z-statistic instead of the t-statistic. Since the former is distributed like a Gaussian it may be easier to deal with.
\begin{align}
z_N = \frac{\hat{p} - \hat{q}}{\sqrt{\hat{r}(1-\hat{r})\left(1/n + 1/m\right)}} \sim \Gaussian(0, 1)
\end{align}
\section{Appendix}

>>>>

By the definition of the chi-squared distribution with one degree of freedom. To reject the null hypothesis $H_0$ at the $100(1-\alpha)\%$ confidence level, we require that $P(T_n \geq t_n | H_0) < \alpha$. In other words, a low probability of observing this result under the null hypothesis.

@@

We now complicate things by introducing a second coin, with throws denoted by $\{Y_i\}_{i=1}^{m}$ where we assume each throw is i.i.d Bernoulli with parameter $q$ ($Y_i \sim Bern(q)$). Note that the population sizes $n$ and $m$ may be different. We define $L \coloneqq \sum_{i=1}^{m} Y_i$ which is the analogue of $K$. We now set up our hypotheses to be:
%
\begin{align*}
H_0:& \quad p = q \\
H_1:& \quad p \neq q
\end{align*}
%
Proceeding as before we can derive a formula for the test statistic. This time we denote the test statistic by $t_N$ where $N \coloneqq n + m$.
%
\begin{align*}
t_N &\coloneqq \log \frac{\lik(H_1)}{\lik(H_0)}
= \log \frac{\max_{p,q} P(K=k|p) P(L=l|q)}{\max_p P(K=k|p) P(L=l|q=p)} \\
&= \log \frac
{\binom{n}{k} \hat{p}^k (1 - \hat{p})^{n-k} \binom{m}{l} \hat{q}^k (1 - \hat{q})^{m-l}}
{\binom{n}{k} \hat{r}^k (1 - \hat{r})^{n-k} \binom{m}{l} \hat{r}^l (1 - \hat{r})^{m-l}} \\
&= \log \frac
{ \hat{p}^k (1 - \hat{p})^{n-k}}
{ \hat{r}^k (1 - \hat{r})^{n-k}}
+ \log \frac{\hat{q}^k (1 - \hat{q})^{m-l}}{ \hat{r}^l (1 - \hat{r})^{m-l}}
\end{align*} 
%
Where, $\hat{p} \coloneqq k/n, \hat{q} \coloneqq l/m, \hat{r} \coloneqq (k+l)/(n+m) = (n\hat{p} + m\hat{q})/(n+m)$. Using the same tricks as before, we can express this in terms of the chi-squared distance between the various parameters:
%
\begin{align*}
t_N &= n \kl\left(Bern(\hat{p}) || Bern(\hat{r})\right)
+ m \kl\left(Bern(\hat{q}) || Bern(\hat{r})\right) \\ 
&\approx \frac{n}{2} \chi^2\left(Bern(\hat{p})||Bern(\hat{r}))\right)
+ \frac{m}{2} \chi^2\left(Bern(\hat{q})||Bern(\hat{r}))\right) \\
&= \frac{1}{2\hat{r}(1-\hat{r})} \left(
n(\hat{p} - \hat{r})^2 + m(\hat{q} - \hat{r})^2
\right) \\
&= \frac{1}{2\hat{r}(1-\hat{r})} \left(
n\left( \frac{m(\hat{p} - \hat{q})}{n+m}\right)^2 + 
m\left( \frac{n(\hat{q} - \hat{p})}{n+m}\right)^2
\right) \\
&= \frac{nm(\hat{p} - \hat{q})^2}{2\hat{r}(1-\hat{r})(n+m)} \\
&= \frac{1}{2} \left( \frac{\left(\hat{p} - \hat{q}\right)}{\sqrt{\hat{r}(1-\hat{r})(1/n+1/m)}}  \right)^2
\end{align*}
%
Under the null hypothesis $H_0$, we require $p=q(=\mu)$; we introduce this third variable $\mu$ to refer to the true mean to avoid ambiguity. Applying the central limit theorem (for sufficiently large $n$ and $m$) and combining Gaussians in the standard way, we have that:
%
\begin{equation*}
\begin{gathered}
\hat{P} \sim \Gaussian \left( \mu, \frac{\mu(1-\mu)}{n} \right) \\
\hat{Q} \sim \Gaussian \left(\mu, \frac{\mu(1-\mu)}{m} \right) \\
\hat{R} \sim \Gaussian \left(\mu, \frac{\mu(1-\mu)}{n+m} \right) \\
\therefore  \sqrt{\frac{nm}{\mu(1-\mu)(n+m)}} (\hat{P} - \hat{Q}) = Z \sim \Gaussian \left(0, 1 \right) \\
\delta \hat{R} \coloneqq \hat{R} - \mu \sim \Gaussian \left( 0, \frac{\mu(1-\mu)}{n+m} \right)
\end{gathered}
\end{equation*}
%
We almost have $T_n$ we just need to demonstrate that $\hat{R}(1-\hat{R})$ is sufficiently close to $\mu(1-\mu)$ for our purposes.
%
\begin{align*}
\hat{R}(1-\hat{R}) &= (\mu + \delta\hat{R})(1-(\mu + \delta \hat{R})) \\
&= \mu(1-\mu) + O(\delta\hat{R}) \\
\therefore \frac{1}{\hat{R}(1-\hat{R})} &= \frac{1}{\mu(1-\mu)} \left(\frac{1}{1+O(\delta\hat{R})}\right) \\
&= \frac{1}{\mu(1-\mu)} (1 + O(\delta\hat{R})) \\
&\approx \frac{1}{\mu(1-\mu)}
\end{align*}
%
We can neglect the terms of order $\delta\hat{R}$ and higher powers, as it is zero mean and for sufficiently large $n+m$ the variance approaches 0. Therefore, we have the desired expression for $T_N$.
%
\begin{equation}
T_N \approx \frac{1}{2} \left(
\sqrt{\frac{nm}{\mu(1-\mu)(n+m)}}
(\hat{P} - \hat{Q})
\right)^2 = \frac{1}{2} Z^2 \sim \frac{1}{2} \chi^2_1
\end{equation}
%
For this test we can also use the z-statistic instead of the t-statistic. Since the former is distributed like a Gaussian it may be easier to deal with.
\begin{align}
z_N = \frac{\hat{p} - \hat{q}}{\sqrt{\hat{r}(1-\hat{r})\left(1/n + 1/m\right)}} \sim \Gaussian(0, 1)
\end{align}

\section{ABP}

ABP(r, T) on a graph $G$ with vertex set $V = V(G)$ and edge set $E = E(G)$
\begin{enumerate}
	\item Initialise messages for $(v, v')$:
		\subitem $y^{(0)}_{v' \rightarrow v} \leftarrow \Gaussian(0, 1)$
	\item Iterate for $1 \leq t \leq T$ and for $(v, v') \in E$:
		\subitem compute average $s^{(t-1)} \leftarrow \frac{1}{2|E|} \sum_{(v, v') \in E} y^{(t-1)}_{v' \rightarrow v}$
		\subitem recentre messages $z^{(t-1)}_{v' \rightarrow v} \leftarrow y^{(t-1)}_{v' \rightarrow v} - s^{(t-1)}$ 
		\subitem sum incoming $y^{(t)}_{v' \rightarrow v} \leftarrow \sum_{(v', v'') \in E \setminus \{v\}}z^{(t-1)}_{v'' \rightarrow v'}$
		\subitem if $(v''' \rightarrow v \rightarrow v')$ on cycle of length $r' \leq r$ then correct:
			\subsubitem $y^{(t)}_{v' \rightarrow v} \leftarrow y^{(t)}_{v' \rightarrow v} - \sum_{(v, v'') \in E \setminus \{v', v'''\}}z^{(t-r')}_{v'' \rightarrow v'}$
	\item Assignment, for all $v \in V$:
		\subitem Sum incoming $y_v^{(T)} = \sum_{(v, v') \in E} y^{(t)}_{v' \rightarrow v}$
		\subitem Assign labels $\sigma_v = 1$ if $y_v^{(T)} > 0$ and $0$ otherwise
\end{enumerate}
\section{Appendix}
